\chapter[Introduction]{Introduction}

\chapterinitial{T}{hank} you for starting to read this book. This book aims to
bring together two fascinating topics:

\begin{itemize}
\item
  Problems that can be solved using mathematics;
\item
  Software that is free to use and change.
\end{itemize}

What we mean by both of those things will become clear through reading
this chapter and the rest of the book.

\section{Who is this book for?}\label{sec:who-is-this-book-for}

Anyone who is interested in using mathematics and computers to solve
problems will hopefully find this book helpful.

%TODO Include VENN Diagram

If you are a student of a mathematical discipline, a graduate student of
a subject like operational research, a hobbyist who enjoys solving the
travelling salesman problem or even if you get paid to do this stuff:
this book is for you. We will introduce you to the world of open source
software that allows you to do all these things freely.

If you are a student learning to write code, a graduate student using
databases for their research, an enthusiast who programmes applications
to help coordinate the neighbourhood watch, or even if you get paid to
write software: this book is for you. We will introduce you to a world
of problems that can be solved using your skill sets.

It would be helpful for the reader of this book to:

\begin{itemize}
\item
  Have access to a computer and be able to connect to the internet (at
  least once) to be able to download the relevant software.
\item
  Be prepared to read some mathematics. Technically you do not need to
  understand the specific mathematics to be able to use the tools in this book.
  The topics covered use some algebra, calculus and probability.
\end{itemize}

\section{What do we mean by applied mathematics?}\label{sec:what-do-we-mean-by-applied-mathematics}

We consider this book to be a book on applied mathematics. This is not
however a universal term, for some applied mathematics is the study of
mechanics and involves modelling projectiles being fired out of canons.
We will use the term a bit more freely here and mean any type of real
world problem that can be tackled using mathematical tools. This is
sometimes referred to as operational research, operations research,
mathematical modelling or indeed just mathematics.

One of the authors, Vince, used mathematics to plan the sitting plan at
his wedding. Using a particular area of mathematics call graph theory he
was able to ensure that everyone sat next to someone they liked and/or
knew.

The other author, Geraint, used mathematics to find the best team of
Pokemon. Using an area of mathematics call linear programming which is
based on linear algebra he was able to find the best makeup of pokemon.

Here, applied mathematics is the type of mathematics that helps us
answer questions that the real world asks.

\section{What is open source
software}\label{sec:what-is-open-source-software}

Strictly speaking open source software is software with source code that
anyone can read, modify and improve. In practice this means that you do
not need to pay to use it which is often one of the first attractions.
This financial aspect can also be one of the reasons that someone will
not use a particular piece of software due to a confusion between cost
and value: if something is free is it really going to be any good?

In practice open source software is used all of the world and powers
some of the most important infrastructure around. For example, one
should never use any cryptographic software that is not open source: if
you cannot open up and read things than you should not trust it (this is
indeed why most cryptographic systems used are open source).
% TODO Add a reference to back up this claim.

Today, open source software is a lot more than a licensing agreement:
it is a community of practice. Bugs are fixed faster, research is
implemented immediately and knowledge is spread more widely thanks to
open source software. Bugs are fixed faster because anyone can read and
inspect the source code. Most open source software projects also have a
clear mechanisms for communicating with the developers and even
reviewing and accepting code contributions from the general public.
Research is implemented immediately because when new algorithms are
discovered they are often added directly to the software by the
researchers who found them. This all contributes to the spread of
knowledge: open source software is the modern should of giants that we
all stand on.

Open source software is software that, like scientific knowledge is not
restricted in its use.

\section{How to get the most out of this
book}\label{sec:how-to-get-the-most-out-of-this-book}

The book itself is open source. You can find the source files for this
book online at \texttt{github.com/drvinceknight/ampwoss}. There will will
also find a number of \emph{Jupyter notebooks} and \emph{R markdown
files} that include code snippets that let you follow along.

We feel that you can choose to read the book from cover to cover,
writing out the code examples as you go; or it could also be used as a
reference text when faced with particular problem and wanting to know
where to start.

The book is made up of 10 chapters that are paired in two 4 parts. Each
part corresponds to a particular area of mathematics, for example
``Emergent Behaviour''. Two chapters are paired together for each
chapter, usually these two chapters correspond to the same area of
mathematics but from a slightly different scale that correspond to
different ways of tackling the problem.

Every chapter has the following structure:

\begin{enumerate}
\item
  Introduction - a brief overview of a given problem type. Here we will
  describe the problem at hand in general terms.
\item
  An Example problem. This will provide a tangible example problem that
  offers the reader some intuition for the rest of the discussion.
\item
  Solving with Python. We will describe the mathematical tools available
  to us in a programming language called Python to solve the problem.
\item
  Solving with R. Here we will do the same with the R programming
  language.
\item
  Brief theoretic background with pointers to reference texts. Some
  readers might like to delve in to the mathematics of the problem a bit
  further, we will include those details here.
\item
  Examples of research using these methods. Finally, some readers might
  even be interested in finding out a bit more of what mathematicians
  are doing on these problems. Often this will include some descriptions
  of the problem considered but perhaps at a much larger scale than the
  one presented in the example.
\end{enumerate}

For a given reader, not all sections of a chapter will be of interest.
Perhaps a reader is only interested in R and finding out more about the
research. Please do take from the book what you find useful.

\section{How code is written in this book}\label{sec:how-to-get-the-most-out-of-this-book}

Throughout this book, there are going to be various pieces of code written. Code
is a series of instructions that usually give some sort of output when submitted
to a computer.

This book will show both the set of instructions (referred to as the input) and
the output.

You will see Python input as follows:

\begin{py-in}
print(2 + 2)
\end{py-in}

and you will see Python output as follows:

\begin{py-out}
4
\end{py-out}

You will see R input as follows:

\begin{R-in}
print(2 + 2)
\end{R-in}

and you will see R output as follows:

\begin{R-out}
4
\end{R-out}

As well as this, a continuous line numbering across all code sections is used so
that if the reader needs to refer to a given set of input or output this can be
done.

The code itself is written using 3 principles:

\begin{itemize}
        \item Modularity: code is written as a
              series of smaller sections of code. These correspond to smaller,
              simpler, individual tasks (modules) that can be used together to carry out
              a particular larger task.
        \item Documentation: readable variable names as
              well as
              text describing the functionality of each module of code are used
              throughout. This ensures that code is not only usable but also
              understandable.
          \item Tests: there are places where each module of code is
              used independently to check the output. This can be thought of as a
              test of functionality which readers can use to check they are
              getting expected results.
\end{itemize}

These are best practice principles in research software development % TODO Add a footnote reference
that ensure usable, reproducible and reliable code.
Interested readers might want to see Figure~\ref{fig:rsd-triangle} which shows
how the 3 principles interact with each other.

% TODO Add rsd triangle diagram.
